---
title: "Module1"
author: "Rohit Kosamkar"
date: "2024-02-27"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Set Source : https://data.cdc.gov/NCHS/Weekly-Provisional-Counts-of-Deaths-by-State-and-S/muzy-jte6/about_data

Data Set Description: Provisional counts of deaths by the week the deaths occurred, by state of occurrence, and by select underlying causes of death for 2020-2023. The dataset also includes weekly provisional counts of death for COVID-19, coded to ICD-10 code U07.1 as an underlying or multiple cause of death.

Time period covered by the data: 2020-2023

Data Size: Rows:10476 Columns:35

Important Columns Description:

"Jurisdiction"            : Refers to the geographical region or jurisdiction where the data was collected.
"septicemia"              : Refers to the number of deaths attributed to septicemia.
"malignant_neoplasms"     : Indicates the number of deaths caused by malignant neoplasms (cancerous tumors).
"diabetes_mellitus"       : Represents the number of deaths due to diabetes mellitus.
"alzheimer_disease"       : Refers to the number of deaths attributed to Alzheimer's disease.
"influenza_and_pneumonia" : Indicates the number of deaths from influenza and pneumonia.
"chronic_respiratory"     : Represents the number of deaths caused by chronic respiratory diseases.
"lab_findings"            : Represents the number of deaths due to abnormalities found in laboratory tests.
"heart_disease"           : Indicates the number of deaths caused by heart disease.
"cerebrovascular_diseases": Refers to the number of deaths attributed to cerebrovascular diseases (related to blood vessels in the brain.
"covid_19_MCD"            : Represents the number of deaths with COVID-19 mentioned as the multiple cause of death (MCD).
"covid_19_UCD"            : Indicates the number of deaths with COVID-19 mentioned as the underlying cause of death (UCD).


Importing Important Libraries

```{r}
library(dplyr)
library(ggplot2)
library(plotly)
library(janitor)
library(zoo)
library(gridExtra)
library(tidyr)
```

```{r}
print(getwd())
setwd("C:/Users/sapna/Downloads")
df <-  read.csv("Counts_of_Deaths.csv")
head(df)

```
Let's check if any NA value present in data set

```{r}
colSums(is.na(df))
```
Let's have a look at column names
```{r}
names(df)
```
Let's drop columns "flag_allcause" and "flag_natcause" contains NA values.
columns [23:35] are empty columns which I will drop for further analysis. 
Let' consider important columns for analysis using subset function
```{r}
df <- subset(df,select= c(2:20))
```
let's clean the columns using janitor's clean_names function
```{r}
names(clean_names(df))
```
Let's rename all the columns as most names of columns are too lengthy
```{r}
names(df) <- c("Jurisdiction", "Year", "Week", "Week_EDT","all_cause", "natural_cause",
               "septicemia","malignant_neoplasms","diabetes_mellitus","alzheimer_disease","influenza_and_pneumonia",
               "chronic_respiratory","other_respiratory", "nephritis_nephrotic","lab_findings","heart_disease",
               "cerebrovascular_diseases","covid_19_MCD", "covid_19_UCD")
names(df)
```
Let's consider data from year 2020-2022 as 2023 not all week's data present and also in Jurisdiction column United States contains  summarize data which is of no use so will drop it and 2 columns all_cause & natural_cause contains total of all columns which I'll drop for further analysis
```{r}
df_sub <- df %>% filter(df$Year< 2023 & df$Jurisdiction!='United States') %>% select(-all_cause, -natural_cause)
names(df_sub)
```
Let's fill NA values with 0, indicating that no deaths occurred due to a particular disease.

```{r}
df_sub[,c(5:17)] <- na.fill(df_sub[, c(5:17)], 0)
colSums(is.na(df_sub))
```
adding Covid related deaths to single column i.e. Multiple cause of death + Underlying cause of death

```{r}
df_sub$Covid_19 <- df_sub$covid_19_MCD + df_sub$covid_19_UCD
df_sub <- subset(df_sub, select = -c(covid_19_MCD, covid_19_UCD))
```
Let's check data types of columns
```{r}
str(df_sub)
```

Summarize year wise data to get total deaths by each disease

```{r}
df_yearwise <- df_sub %>%
  group_by(Year) %>%
  summarise(
    across(septicemia:Covid_19, sum)
  )
```

Creating dataframes for 3-years 2020, 2021, 2022
```{r}
df_yearwise
```

Let's write a function to analyze and visualize total deaths by Disease every year

```{r}
plot_disease_deaths <- function(df, year){
  df <- df %>% pivot_longer(cols= -Year, names_to = "Disease", values_to = "Deaths") %>% 
    arrange(desc(Deaths)) %>% 
    mutate(Disease = factor(Disease, levels = unique(Disease)))
  
  ggplot(df, aes(x= reorder(Disease, -Deaths), y= Deaths/1e6))+
    geom_bar(stat= "identity", fill= "skyblue") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14),
          plot.title = element_text(size = 16)) +
    labs(title = paste("Death by Disease in", year),x = "Disease", y= "Deaths (Millions)" )
}

```


```{r}
plot_disease_deaths(df_yearwise[1,], 2020)
```


```{r}
plot_disease_deaths(df_yearwise[2,], 2021)
```


```{r}
plot_disease_deaths(df_yearwise[3,], 2022)
```
Let's see what all values are there in Jurisdiction

```{r}
unique(df_sub$Jurisdiction)
#names(df_sub)
```
Let's write a function to analyze total deaths year wise for all specified diseases for top 10 states

```{r}
plot_state_deaths <- function(df,year){
  main_df <- df %>% filter(Year==year) %>% group_by(Jurisdiction) %>% 
    summarise(deaths= sum(c(septicemia,malignant_neoplasms,diabetes_mellitus,
                            alzheimer_disease,influenza_and_pneumonia,
                            chronic_respiratory,other_respiratory,
                            nephritis_nephrotic,lab_findings,heart_disease,
                            cerebrovascular_diseases,cerebrovascular_diseases))) %>% 
    arrange(desc(deaths))  
  
  ggplot(main_df %>% head(10), aes(x = reorder(Jurisdiction, -deaths), y = deaths))+
  scale_y_continuous(labels = scales::comma)+
  geom_bar(stat= "identity", fill= "pink") +
  geom_hline(yintercept = mean(main_df$deaths), linetype = "dashed", color = "blue",show.legend = TRUE) +  
  geom_text(aes(x = Inf, y = mean(main_df$deaths), 
                label = paste("Average:", round(mean(main_df$deaths), 0))), vjust = -1, hjust = 1, color = "blue") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14),
        plot.title = element_text(size = 16)) +
  labs(title = paste("Top 10 Jurisdiction by total deaths", (year)),x = "Jurisdiction", y= "Deaths" )
}

```
Year 2020
```{r}
plot_state_deaths(df_sub, 2020)
```
Year 2021
```{r}
plot_state_deaths(df_sub, 2021)
```
Year 2022
```{r}
plot_state_deaths(df_sub, 2022)
```
Since Now I found California, Florida, Texas and Pennsylvania has most of the death happened, let's see trends of death week wise for Covid-19, heart disease and Malingnant neoplasms which caused the most deaths among all the diseases

```{r}
trends_top_states_disease <- function(df, year, jurisdiction){
  
df_op <- df %>% filter(Year==year & Jurisdiction==jurisdiction) %>% select(c(Jurisdiction, Year, Week, Covid_19, heart_disease, malignant_neoplasms))

ggplot(df_op, aes(x = Week)) +
  geom_line(aes(y = Covid_19, color = "Covid_19"), linewidth = 1) +
  geom_line(aes(y = heart_disease, color = "Heart Disease"), linewidth = 1) +
  geom_line(aes(y = malignant_neoplasms, color = "Malignant Neoplasms"), linewidth = 1) +
  scale_color_manual(values = c("Covid_19" = "blue", "Heart Disease" = "red", "Malignant Neoplasms" = "green")) +
  theme_minimal() +
  labs(title = paste("Weekly Trends of Diseases",jurisdiction,  (year)),
       x = "Week",
       y = "Number of Cases",
       color = "Disease") +
  theme(legend.position = "right",
        legend.text = element_text(size = 12),
      legend.title = element_text(size = 13), 
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13),
        axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        plot.title = element_text(size = 15)) +
  theme(text = element_text(size = 13)) +
  scale_x_continuous(expand = c(0, 0))
}
```

For the year 2020
```{r fig.height=10, fig.width=15}
calf_20 <- trends_top_states_disease(df_sub, 2020,"California")
flor_20 <- trends_top_states_disease(df_sub, 2020,"Florida")
texas_20 <- trends_top_states_disease(df_sub, 2020,"Texas")
penn_20 <- trends_top_states_disease(df_sub, 2020, "Pennsylvania")
grid.arrange(calf_20, flor_20,texas_20,penn_20, ncol =2)
```

For the year 2021
```{r fig.height=10, fig.width=15}
calf_21 <- trends_top_states_disease(df_sub, 2021,"California")
flor_21 <- trends_top_states_disease(df_sub, 2021,"Florida")
texas_21 <- trends_top_states_disease(df_sub, 2021,"Texas")
penn_21 <- trends_top_states_disease(df_sub, 2021, "Pennsylvania")
grid.arrange(calf_21, flor_21,texas_21,penn_21, ncol =2)
```

For the year 2022
```{r fig.height=10, fig.width=15}
calf_22 <- trends_top_states_disease(df_sub, 2022,"California")
flor_22 <- trends_top_states_disease(df_sub, 2022,"Florida")
texas_22 <- trends_top_states_disease(df_sub, 2022,"Texas")
penn_22 <- trends_top_states_disease(df_sub, 2022, "Pennsylvania")
grid.arrange(calf_22, flor_22,texas_22,penn_22, ncol =2)
```
Let's see Bottom 10 Jurisdiction by total deaths year wise

```{r}
low_deaths_df <- df_sub %>%
    filter(Year %in% c(2020, 2021, 2022)) %>%
    group_by(Year,Jurisdiction) %>%
    summarise(deaths = sum(c(septicemia, malignant_neoplasms, diabetes_mellitus,
                             alzheimer_disease, influenza_and_pneumonia,
                             chronic_respiratory, other_respiratory,
                             nephritis_nephrotic, lab_findings, heart_disease,
                             cerebrovascular_diseases, cerebrovascular_diseases))) %>%
    arrange(Year,deaths) %>% 
    slice(1:10)
```


```{r}
ggplot(low_deaths_df, aes(x = Jurisdiction, y = deaths, fill = as.factor(Year))) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14),
          plot.title = element_text(size = 16)) +
    labs(title = "Lower Total Deaths Counts by Jurisdiction (2020, 2021, 2022)",
         x = "Jurisdiction", y = "Total Deaths",
         fill = "Year") +
    scale_fill_manual(values = c("2020" = "orange", "2021" = "violet", "2022" = "blue")) +
    guides(fill = guide_legend(title = "Year"))
```





